### 2025-03-19 PAPER_CODE.qmd
Add code for analysis with ZIPLN

---
title: "Species association networks unveil scale effects in community assemblage mechanisms."
author: "Jeanne Tous, Julien Chiquet"
format: html
editor: visual
---

## Useful libraries & functions

```{r}
library(dplyr)
# library(PLNmodels)
library(igraph)
```

```{r}
plot_network = function(net             ,
                         type            ,
                         output          ,
                         edge.color      = c("#F8766D", "#00BFC4"),
                         remove.isolated = FALSE,
                         node.labels     = NULL,
                         layout          = layout_in_circle,
                         plot = TRUE) {

  if (output == "igraph") {

    G <-  graph_from_adjacency_matrix(net, mode = "undirected", weighted = TRUE, diag = FALSE)

    if (!is.null(node.labels)) {
      igraph::V(G)$label <- node.labels
    } else {
      igraph::V(G)$label <- colnames(net)
    }
    ## Nice nodes
    V.deg <- degree(G)/sum(degree(G))
    igraph::V(G)$label.cex <- V.deg / max(V.deg) + .5
    igraph::V(G)$size <- V.deg * 100
    igraph::V(G)$label.color <- rgb(0, 0, .2, .8)
    igraph::V(G)$frame.color <- NA
    ## Nice edges
    igraph::E(G)$color <- ifelse(igraph::E(G)$weight > 0, edge.color[1], edge.color[2])
    if (type == "support")
      igraph::E(G)$width <- abs(igraph::E(G)$weight)
    else
      igraph::E(G)$width <- 15*abs(igraph::E(G)$weight)

    if (remove.isolated) {
      G <- delete.vertices(G, which(degree(G) == 0))
    }
    if (plot) plot(G, layout = layout)
  }
  if (output == "corrplot") {
    if (plot) {
      if (ncol(net) > 100)
        colnames(net) <- rownames(net) <- rep(" ", ncol(net))
      G <- net
      diag(net) <- 0
      corrplot(as.matrix(net), method = "color", is.corr = FALSE, tl.pos = "td", cl.pos = "n", tl.cex = 0.5, type = "upper")
    } else  {
      G <- net
    }
  }
  invisible(G)
}

```

## Loading & formatting data

### Regional dataset

```{r}
load("REGIONAL_DATASET_COUNTS.Rdata")
load("REGIONAL_DATASET_COVARIATES.Rdata")
```

```{r}
regional_data <- prepare_data(counts     = REGIONAL_DATASET_COUNTS,
                              covariates = REGIONAL_DATASET_COVARIATES)
```

### Northern Range dataset

```{r}
load("NORTHERN_RANGE_DATASET_COUNTS.Rdata")
load("NORTHERN_RANGE_DATASET_COVARIATES.Rdata")
```

```{r}
northern_range_data <- prepare_data(counts     = NORTHERN_RANGE_DATASET_COUNTS,
                                    covariates = NORTHERN_RANGE_DATASET_COVARIATES)
```

## Running the ZIPLN-network model

We fix the seed for reproducibility.

```{r}
set.seed(1)
```

### Regional dataset

```{r}
regional_model <- ZIPLNnetwork(Abundance ~ 1 + LONGLAT, data = regional_data, control = ZIPLNnetwork_param(min_ratio = 0.001,  penalize_diagonal = FALSE), zi = "col")

```

We select a penalty using stability selection process StARS. As it stability selection requires a long time to run, we memorize the corresponding penalty and later use it directly to access the model selected with StARS.

```{r}
# regional_model_StARS <- regional_model$getBestModel("StARS")
regional_model_StARS <- regional_model$getModel(0.06881528)  # Avec le code git
# regional_model_StARS <- regional_model$getModel(0.4222431) # Avec la librairie
```

### Northern Range dataset

```{r}
northern_range_model <- ZIPLNnetwork(Abundance ~ 1 + LONGLAT, data = northern_range_data, control = ZIPLNnetwork_param(min_ratio = 0.001,  penalize_diagonal = FALSE), zi = "col")

```

We select a penalty using stability selection process StARS. As it stability selection requires a long time to run, we memorize the corresponding penalty and later use it directly to access the model selected with StARS.

```{r}

# northern_range_model_StARS <- northern_range_model$getBestModel("StARS")
northern_range_model_StARS <- northern_range_model$getModel(0.09219056)
```

## Resulting networks

We want to place species the datasets share in the same order for readibility.

```{r}
regional_species       <- colnames(regional_data$Abundance)
northern_range_species <- colnames(northern_range_data$Abundance)
intersecting_species   <- intersect(northern_range_species, regional_species)
```

### Regional network

Sorting species names.

```{r}
regional_sorted_species <- c(intersecting_species, regional_species[! regional_species %in% intersecting_species])
```

Extracting the association network from the model.

```{r}
regional_network        <- as.matrix(regional_model_StARS$latent_network("partial_cor"))
regional_sorted_indices <- match(regional_sorted_species, colnames(regional_network))
regional_network        <- regional_network[regional_sorted_indices, regional_sorted_indices] 
```

```{r}
plot_network(regional_network,
             type            = "partial_cor",
             output          = "igraph",
             edge.color      = c("#F8766D", "#00BFC4"),
             remove.isolated = FALSE,
             node.labels     = NULL,
             layout          = layout_in_circle,
             plot            = TRUE)
```

### Northern Range network

Sorting species names.

```{r}
northern_range_sorted_species <- c(intersecting_species, northern_range_species[! northern_range_species %in% intersecting_species])
```

```{r}
northern_range_network        <- as.matrix(northern_range_model_StARS$latent_network("partial_cor"))
northern_range_sorted_indices <- match(northern_range_sorted_species, colnames(northern_range_network))
northern_range_network        <- northern_range_network[northern_range_sorted_indices, northern_range_sorted_indices] 
```

```{r}
plot_network(northern_range_network,
             type            = "partial_cor",
             output          = "igraph",
             edge.color      = c("#F8766D", "#00BFC4"),
             remove.isolated = FALSE,
             node.labels     = NULL,
             layout          = layout_in_circle,
             plot            = TRUE)
```

### 2025-03-21 PAPER_CODE.qmd
named code chunks and added html results from qmd file

@@
 
-```{r}
+PLNmodels version should be 1.2.1 or higher.
+
+```{r, loading libraries}
 library(dplyr)
-# library(PLNmodels)
+library(PLNmodels)
 library(igraph)
@@
 
-```{r}
+```{r, function to plot networks}
 plot_network = function(net             ,
                         type            ,
-                         output          ,
                         edge.color      = c("#F8766D", "#00BFC4"),
@@
                         node.labels     = NULL,
-                         layout          = layout_in_circle,
-                         plot = TRUE) {
-
-  if (output == "igraph") {
-
+                        layout          = layout_in_circle) {
   G <-  graph_from_adjacency_matrix(net, mode = "undirected", weighted = TRUE, diag = FALSE)
@@
   }
-    if (plot) plot(G, layout = layout)
-  }
-  if (output == "corrplot") {
-    if (plot) {
-      if (ncol(net) > 100)
-        colnames(net) <- rownames(net) <- rep(" ", ncol(net))
-      G <- net
-      diag(net) <- 0
-      corrplot(as.matrix(net), method = "color", is.corr = FALSE, tl.pos = "td", cl.pos = "n", tl.cex = 0.5, type = "upper")
-    } else  {
-      G <- net
-    }
-  }
+  plot(G, layout = layout)
   invisible(G)
@@
 
-```{r}
+```{r, loading regional dataset}
 load("REGIONAL_DATASET_COUNTS.Rdata")
@@
 
-```{r}
+```{r, formatting regional data for PLNmodels}
 regional_data <- prepare_data(counts     = REGIONAL_DATASET_COUNTS,
@@
 
-```{r}
+```{r, loading Northern Range dataset}
 load("NORTHERN_RANGE_DATASET_COUNTS.Rdata")
@@
 
-```{r}
+```{r, formatting Northern Range data for PLNmodels}
 northern_range_data <- prepare_data(counts     = NORTHERN_RANGE_DATASET_COUNTS,
@@
 
-```{r}
+```{r, fixing the seed}
 set.seed(1)
@@
 
-```{r}
+```{r, running ZIPLNnetwork on the regional data}
 regional_model <- ZIPLNnetwork(Abundance ~ 1 + LONGLAT, data = regional_data, control = ZIPLNnetwork_param(min_ratio = 0.001,  penalize_diagonal = FALSE), zi = "col")
@@
 
-```{r}
+```{r, fixing regularization penalty with StARS - regional}
 # regional_model_StARS <- regional_model$getBestModel("StARS")
-regional_model_StARS <- regional_model$getModel(0.06881528)  # Avec le code git
-# regional_model_StARS <- regional_model$getModel(0.4222431) # Avec la librairie
+regional_model_StARS <- regional_model$getModel(0.06950515) 
 ```
@@
 
-```{r}
+```{r, running ZIPLNnetwork on the Northern Range data}
 northern_range_model <- ZIPLNnetwork(Abundance ~ 1 + LONGLAT, data = northern_range_data, control = ZIPLNnetwork_param(min_ratio = 0.001,  penalize_diagonal = FALSE), zi = "col")
@@
 
-```{r}
+```{r, fixing regularization penalty with StARS - Northern Range}
 
 # northern_range_model_StARS <- northern_range_model$getBestModel("StARS")
-northern_range_model_StARS <- northern_range_model$getModel(0.09219056)
+northern_range_model_StARS <- northern_range_model$getModel(0.08646393)
 ```
@@
 
-```{r}
+```{r, retrieving the lists of species to sort them}
 regional_species       <- colnames(regional_data$Abundance)
@@
 
-```{r}
+```{r, sorting regional species}
 regional_sorted_species <- c(intersecting_species, regional_species[! regional_species %in% intersecting_species])
@@
 
-```{r}
+```{r, sorting regional network}
 regional_network        <- as.matrix(regional_model_StARS$latent_network("partial_cor"))
@@
 
-```{r}
+```{r, plotting regional network}
 plot_network(regional_network,
              type            = "partial_cor",
-             output          = "igraph",
              edge.color      = c("#F8766D", "#00BFC4"),
@@
              node.labels     = NULL,
-             layout          = layout_in_circle,
-             plot            = TRUE)
+             layout          = layout_in_circle)
 ```
@@
 
-```{r}
+```{r,  sorting Northern Range species}
 northern_range_sorted_species <- c(intersecting_species, northern_range_species[! northern_range_species %in% intersecting_species])
@@
 
-```{r}
+```{r, sorting Northern Range network}
 northern_range_network        <- as.matrix(northern_range_model_StARS$latent_network("partial_cor"))
@@
 
-```{r}
+```{r, plotting Northern Range network}
 plot_network(northern_range_network,
              type            = "partial_cor",
-             output          = "igraph",
              edge.color      = c("#F8766D", "#00BFC4"),

### 2025-04-18 methodological_analysis.qmd
Add files via upload

---
title: "ZIPLN-network analysis"
format: html
editor: visual
---

# General questions

We aim at using the ZIPLN-network framework to answer the following questions:

1.  Does accounting for the zero-inflation in the data improve the results of the model ? How so ?

2.  Are the two datasets' distributions significantly different ?

3.  How do the environmental covariates influence the results ? Which ones should we include ?

4.  How robust are the different association to changes in the model (inclusion of different covariates, zero-inflation modelling...) ?

5.  Does including sampling months / years change the results for the Northern Range dataset ?

# Useful libraries & functions

```{r loading_libraries}
library(dplyr)
library(PLNmodels)
library(igraph)
library(FactoMineR)
library(ggplot2)
library(Metrics)
library(Hmsc)
library(corrplot)
set.seed(1)
```

```{r robustness}

robustness = function(models, models_names = NULL,
                      matrices_input = FALSE){
  N <- length(models) ; res <- matrix(rep(0, N * N), nrow = N) ; diag(res) <- 1
  for(i in 1:(N-1)){
    if(!matrices_input){
      mod1 <- models[[i]] ; net1   <- as.matrix(mod1$latent_network("partial_cor"))
      diag(net1) <- 0
    }else{net1 <- models[[i]]}
    count1 <- sum(net1 != 0)
    for(j in ((i + 1):N)){
      if(!matrices_input){
        mod2 <- models[[j]] ; net2   <- as.matrix(mod2$latent_network("partial_cor"))
        diag(net2) <- 0
      }else{net2 <- models[[j]]}
      count2 <- sum(net2 != 0)
      intersection <- sum(sign(net1) * sign(net2) == 1)
      res[i, j]    <- round(intersection / count1, 2)
      res[j, i]    <- round(intersection / count2, 2)
    }
  }
  if(!is.null(models_names)){
    rownames(res) <- models_names ; colnames(res) <- models_names
  }
  cat("M[i,j] gives the proportion of interactions contained in model i that are also in model j. \n")
  return(res)
}
```

```{r plot-network}
plot_network = function(net             ,
                        type = "partial_cor",            
                        edge.color      = c("#F8766D", "#00BFC4"),
                        remove.isolated = FALSE,
                        node.labels     = NULL,
                        layout          = layout_in_circle) {
  G <-  graph_from_adjacency_matrix(net, mode = "undirected", weighted = TRUE, diag = FALSE)

  if (!is.null(node.labels)) {
    igraph::V(G)$label <- node.labels
  } else {
    igraph::V(G)$label <- colnames(net)
  }
  ## Nice nodes
  V.deg <- degree(G)/sum(degree(G))
  igraph::V(G)$label.cex <- V.deg / max(V.deg) + .5
  igraph::V(G)$size <- V.deg * 100
  igraph::V(G)$label.color <- rgb(0, 0, .2, .8)
  igraph::V(G)$frame.color <- NA
  ## Nice edges
  igraph::E(G)$color <- ifelse(igraph::E(G)$weight > 0, edge.color[1], edge.color[2])
  if (type == "support")
    igraph::E(G)$width <- abs(igraph::E(G)$weight)
  else
    igraph::E(G)$width <- 15*abs(igraph::E(G)$weight)

  if (remove.isolated) {
    G <- delete.vertices(G, which(degree(G) == 0))
  }
  plot(G, layout = layout, font.size = 8)
  invisible(G)
}

```

# Data importation

### Regional dataset

```{r, loading regional dataset}
load("REGIONAL_DATASET_COUNTS.Rdata")
load("REGIONAL_DATASET_COVARIATES.Rdata")
```

```{r, formatting regional data for PLNmodels}
regional_data <- prepare_data(counts     = REGIONAL_DATASET_COUNTS,
                              covariates = REGIONAL_DATASET_COVARIATES)
```

### Northern Range dataset

```{r, loading Northern Range dataset}
load("NORTHERN_RANGE_DATASET_COUNTS.Rdata")
load("NORTHERN_RANGE_DATASET_COVARIATES.Rdata")
```

```{r, formatting Northern Range data for PLNmodels}
northern_range_data <- prepare_data(counts     = NORTHERN_RANGE_DATASET_COUNTS,
                                    covariates = NORTHERN_RANGE_DATASET_COVARIATES)
```

# Data visualization

We use log-scale heatmaps to visualize the data. \## Regional dataset

```{r regional_heatmap}
pheatmap::pheatmap(log(1 + regional_data$Abundance),
                   treeheight_row = 0, treeheight_col = 0, show_rownames = FALSE)
```

## Northern Range dataset

```{r NR_heatmap}
pheatmap::pheatmap(log(1 + northern_range_data$Abundance),
                   treeheight_row = 0, treeheight_col = 0, show_rownames = FALSE)
```

# Preparing the covariates

Both datasets have the following environmental covariates in common: width, depth, temperature, oxygen level and pH. Given the results, we choose to keep the first three principal components.

## PCA on the covariates

### Regional dataset

```{r standardizing_covariates_regional}
regional_covar <- regional_data %>% select(c("WIDTHavg", "DEPTHavg",
                                             "TEMPavg", "OXYGENavg",
                                             "pHaverage"))
regional_covar <- as.data.frame(scale(regional_covar))
```

```{r pca_regional}
pca_regional <- PCA(regional_covar)
print(pca_regional$eig)
```

Given the PCA results, the site indexed 40 seems to be an outlier. We remove it before further analysis.

```{r}
REGIONAL_DATASET_COUNTS     <- REGIONAL_DATASET_COUNTS[-c(40),]
REGIONAL_DATASET_COVARIATES <- REGIONAL_DATASET_COVARIATES[-c(40),]
regional_data <- prepare_data(counts     = REGIONAL_DATASET_COUNTS,
                              covariates = REGIONAL_DATASET_COVARIATES)
```

```{r}
pca_covar_regional <- data.frame("PC1" = unname(pca_regional$ind$coord[, 1]),
                                 "PC2" = unname(pca_regional$ind$coord[, 2]),
                                 "PC3" = unname(pca_regional$ind$coord[, 3]))
pca_covar_regional <- pca_covar_regional[-c(40),]
rownames(pca_covar_regional) <- NULL
```

```{r}
regional_data_pc <- prepare_data(counts = REGIONAL_DATASET_COUNTS,
                                 covariates = pca_covar_regional)
```

### Northern Range dataset

```{r standardizing_covariates_nr}
northern_range_covar <- northern_range_data %>%
                        select(c("width", "depth", "temperature", "O2", "pH"))
northern_range_covar <- as.data.frame(scale(northern_range_covar))
```

```{r pca_northern_range}
pca_northern_range <- PCA(northern_range_covar)
print(pca_northern_range$eig)
```

```{r}
pca_covar_northern_range <- data.frame("PC1" = unname(pca_northern_range$ind$coord[, 1]),
                                       "PC2" = unname(pca_northern_range$ind$coord[, 2]),
                                       "PC3" = unname(pca_northern_range$ind$coord[, 3]))
rownames(pca_covar_northern_range) <- rownames(NORTHERN_RANGE_DATASET_COUNTS)
```

```{r}
northern_range_data_pc <- prepare_data(counts     = NORTHERN_RANGE_DATASET_COUNTS,
                                       covariates = pca_covar_northern_range)
```

## Trying to get the altitude
```{r}
library(elevatr)
library(sf)
library(terra)
```

```{r}
# point <- st_as_sf(data.frame(x = -61.31968, y = 10.70517),
#                   coords = c("x", "y"), crs = 4326)
# 
# # Get elevation
# elevation <- get_elev_point(point, src = "aws")
# print(elevation)

altitude <- unlist(lapply(1:304, F = function(i){
  point <- st_as_sf(data.frame(x = northern_range_data$LONGITUDE[[i]], y = northern_range_data$LATITUDE[[i]]),
                  coords = c("x", "y"), crs = 4326)
  return(get_elev_point(point, src = "aws")$elevation)
}))

```

```{r}
get_elev_point(locations = mts, prj = ll_prj)
```

# PLN and ZIPLN (no network penalty) to compare the fitting qualities

## PLN - PC1 + PC2

### regional dataset

```{r}
regional_no_network <- PLN(Abundance ~ 1 + PC1 + PC2,
                             data = regional_data_pc)
```

```{r}
Y_real   <- regional_data_pc$Abundance
Y_fitted <- regional_no_network$fitted ; colnames(Y_fitted) <- colnames(Y_real)
regional_rmse <- rmse(Y_real, Y_fitted)
plot(Y_real, Y_fitted) ; print(regional_rmse)  
```

### NR dataset

```{r}
northern_range_no_network <- PLN(Abundance ~ 1 + PC1 + PC2,
                                 data = northern_range_data_pc)
```

```{r}
Y_real   <- northern_range_data_pc$Abundance
Y_fitted <- northern_range_no_network$fitted ; colnames(Y_fitted) <- colnames(Y_real)
northern_range_rmse <- rmse(Y_real, Y_fitted)
plot(Y_real, Y_fitted) ; print(northern_range_rmse)  
```

```{r}
fitting_results <- data.frame("Model" = c("PLN_PC1_PC2"),
                              "regional_RMSE" = c(round(regional_rmse, 3)),
                              "northern_range_RMSE" = c(round(northern_range_rmse, 3)))
```

## ZIPLN - PC1 + PC2

### regional dataset

```{r}
regional_no_network_ZI <- ZIPLN(Abundance ~ 1 + PC1 + PC2 | 1,
                                data = regional_data_pc)
```

```{r}
Y_real   <- regional_data_pc$Abundance
Y_fitted <- regional_no_network_ZI$fitted ; colnames(Y_fitted) <- colnames(Y_real)
regional_rmse <- rmse(Y_real, Y_fitted)
plot(Y_real, Y_fitted) ; print(regional_rmse)  
```

### NR dataset

```{r}
northern_range_no_network_ZI <- ZIPLN(Abundance ~ 1 + PC1 + PC2 | 1,
                                      data = northern_range_data_pc)
```

```{r}
Y_real   <- northern_range_data_pc$Abundance
Y_fitted <- northern_range_no_network_ZI$fitted ; colnames(Y_fitted) <- colnames(Y_real)
northern_range_rmse <- rmse(Y_real, Y_fitted)
plot(Y_real, Y_fitted) ; print(northern_range_rmse)  
```

```{r}
fitting_results <- rbind(fitting_results, c("ZIPLN_PC1_PC2_ZIcol",
                                            round(regional_rmse, 3),
                                            round(northern_range_rmse, 3)))
```

## ZIPLN - Covariates effects in the ZI

### ZI PC1

#### regional dataset

```{r}
regional_no_network_ZI_PC1 <- ZIPLN(Abundance ~ 1 + PC1 + PC2 | 1 + PC1 ,
                                data = regional_data_pc)
```

```{r}
Y_real   <- regional_data_pc$Abundance
Y_fitted <- regional_no_network_ZI$fitted ; colnames(Y_fitted) <- colnames(Y_real)
regional_rmse <- rmse(Y_real, Y_fitted)
plot(Y_real, Y_fitted) ; print(regional_rmse)  
```

#### NR dataset

```{r}
northern_range_no_network_ZI <- ZIPLN(Abundance ~ 1 + PC1 + PC2  | 1 + PC1,
                                      data = northern_range_data_pc)
```

```{r}
Y_real   <- northern_range_data_pc$Abundance
Y_fitted <- northern_range_no_network_ZI$fitted ; colnames(Y_fitted) <- colnames(Y_real)
northern_range_rmse <- rmse(Y_real, Y_fitted)
plot(Y_real, Y_fitted) ; print(northern_range_rmse)  
```

```{r}
fitting_results <- rbind(fitting_results, c("ZIPLN_PC1_PC2_ZIcolPC1",
                                            round(regional_rmse, 3),
                                            round(northern_range_rmse, 3)))
```

### ZI PC2

#### regional dataset

```{r}
regional_no_network_ZI_PC1_PC2 <- ZIPLN(Abundance ~ 1 + PC1 + PC2 | 1 + PC1 + PC2,
                                        data = regional_data_pc)
```

```{r}
Y_real   <- regional_data_pc$Abundance
Y_fitted <- regional_no_network_ZI$fitted ; colnames(Y_fitted) <- colnames(Y_real)
regional_rmse <- rmse(Y_real, Y_fitted)
plot(Y_real, Y_fitted) ; print(regional_rmse)  
```

#### NR dataset

```{r}
# northern_range_no_network_ZI <- ZIPLN(Abundance ~ 1 + PC1 + PC2  | 1 + PC1 + PC2,
#                                       data = northern_range_data_pc)
```

```{r}
# Y_real   <- northern_range_data_pc$Abundance
# Y_fitted <- northern_range_no_network_ZI$fitted ; colnames(Y_fitted) <- colnames(Y_real)
# northern_range_rmse <- rmse(Y_real, Y_fitted)
# plot(Y_real, Y_fitted) ; print(northern_range_rmse)  
northern_range_rmse <- NA
```

```{r}
fitting_results <- rbind(fitting_results, c("ZIPLN_PC1_PC2_ZIcolPC1PC2",
                                            round(regional_rmse, 3),
                                            round(northern_range_rmse, 3)))
```

```{r}
print(fitting_results)
```

# Running PLN-network (no zero-inflation)

### Fitting the model

#### Regional dataset

```{r regional_model_no_ZI}
regional_no_ZI <- PLNnetwork(Abundance ~ 1 + PC1 + PC2,
                             data = regional_data_pc,
                             control = PLNnetwork_param(min_ratio = 0.05,
                                                        penalize_diagonal = FALSE))
```

```{r}
# regional_no_ZI_StARS <- regional_no_ZI$getBestModel("StARS")
regional_no_ZI_StARS <- regional_no_ZI$getModel(1.191822)
```

#### Northern Range dataset

```{r northern_range_model_no_ZI}
northern_range_no_ZI <- PLNnetwork(Abundance ~ 1 + PC1 + PC2,
                                  data = northern_range_data_pc,
                                  control = PLNnetwork_param(min_ratio = 0.05,
                                                             penalize_diagonal = FALSE))
```

```{r}
# northern_range_no_ZI_StARS <- northern_range_no_ZI$getBestModel("StARS")
northern_range_no_ZI_StARS <- northern_range_no_ZI$getModel(0.6133228)
```

### Goodness of fit

#### Regional dataset

```{r regional_fit}
Y_real   <- regional_data_pc$Abundance
Y_fitted <- regional_no_ZI_StARS$fitted ; colnames(Y_fitted) <- colnames(Y_real)
plot(Y_real, Y_fitted)
```

```{r}
print(rmse(Y_real, Y_fitted))
```

Focusing on the species with many 0s.

```{r regional_fit_zeros}
zero_threshold <- median(colSums(Y_real == 0))
zero_species   <- names(which(colSums(Y_real == 0) > zero_threshold & colSums(Y_real) < 400))
Y_zero         <- Y_real[, zero_species] ; Y_fitted_zero <- Y_fitted[, zero_species] 
plot(Y_zero, Y_fitted_zero)
```

```{r}
print(rmse(Y_zero, Y_fitted_zero))
```

#### Northern Range dataset

```{r nr_fit}
Y_real   <- northern_range_data$Abundance
Y_fitted <- northern_range_no_ZI_StARS$fitted ; colnames(Y_fitted) <- colnames(Y_real)
plot(Y_real, Y_fitted)
```

```{r}
print(rmse(Y_real, Y_fitted))
```

Focusing on the species with many 0s.

```{r nr_fit_zeros}
zero_threshold <- median(colSums(Y_real == 0))
zero_species   <- names(which(colSums(Y_real == 0) > zero_threshold & colSums(Y_real) < 400))
Y_zero         <- Y_real[, zero_species] ; Y_fitted_zero <- Y_fitted[, zero_species] 
plot(Y_zero, Y_fitted_zero)
```

```{r}
print(rmse(Y_zero, Y_fitted_zero))
```

### Resulting networks

#### Regional dataset

```{r}
plot(regional_no_ZI_StARS)
```

With a 0 penalty we would have:

```{r}
plot(regional_no_ZI$getModel(0))
```

#### Northern Range dataset

```{r}
plot(northern_range_no_ZI_StARS)
```

With a 0 penalty we would have:

```{r}
plot(northern_range_no_ZI$getModel(0))
```

### Robustness of the networks to the addition of covariates

#### Regional network

```{r}
plot(regional_no_ZI_StARS)
```

```{r}
regional_no_ZI_PC3 <- PLNnetwork(Abundance ~ 1 + PC1 + PC2 + PC3,
                                 data = regional_data_pc,
                                 control = PLNnetwork_param(min_ratio = 0.05,
                                                            penalize_diagonal = FALSE))
# regional_no_ZI_PC3_StARS <- regional_no_ZI_PC3$getBestModel("StARS")
regional_no_ZI_PC3_StARS <- regional_no_ZI_PC3$getModel(1.15029)
plot(regional_no_ZI_PC3_StARS)
```

```{r}
regional_no_ZI_LONGLAT <- PLNnetwork(Abundance ~ 1 + LONGLAT,
                                 data = regional_data,
                                 control = PLNnetwork_param(min_ratio = 0.05,
                                                            penalize_diagonal = FALSE))
# regional_no_ZI_LONGLAT_StARS <- regional_no_ZI_LONGLAT$getBestModel("StARS")
regional_no_ZI_LONGLAT_StARS <- regional_no_ZI_LONGLAT$getModel(1.153061)
plot(regional_no_ZI_LONGLAT_StARS)
```

```{r}
robustness(c(regional_no_ZI_StARS,
             regional_no_ZI_PC3_StARS,
             regional_no_ZI_LONGLAT_StARS),
           c("regional_no_ZI_StARS",
             "regional_no_ZI_PC3_StARS",
             "regional_no_ZI_LONGLAT_StARS" ))
```

#### Northern Range network

```{r}
plot(northern_range_no_ZI_StARS)
```

```{r}
northern_range_no_ZI_PC3 <- PLNnetwork(Abundance ~ 1 + PC1 + PC2 + PC3,
                                       data = northern_range_data_pc,
                                       control = PLNnetwork_param(min_ratio = 0.05,
                                                            penalize_diagonal = FALSE))
# northern_range_no_ZI_PC3_StARS <- northern_range_no_ZI_PC3$getBestModel("StARS")
northern_range_no_ZI_PC3_StARS <- northern_range_no_ZI_PC3$getModel(0.5743538)
plot(northern_range_no_ZI_PC3_StARS)
```

```{r}
northern_range_no_ZI_LONGLAT <- PLNnetwork(Abundance ~ 1 + LONGLAT,
                                 data = northern_range_data,
                                 control = PLNnetwork_param(min_ratio = 0.05,
                                                            penalize_diagonal = FALSE))
# northern_range_no_ZI_LONGLAT_StARS <- northern_range_no_ZI_LONGLAT$getBestModel("StARS")
northern_range_no_ZI_LONGLAT_StARS <- northern_range_no_ZI_LONGLAT$getModel(0.6260702)
plot(northern_range_no_ZI_LONGLAT_StARS)
```

```{r}
robustness(c(northern_range_no_ZI_StARS,
             northern_range_no_ZI_PC3_StARS,
             northern_range_no_ZI_LONGLAT_StARS),
           c("northern_range_no_ZI_StARS",
             "northern_range_no_ZI_PC3_StARS",
             "northern_range_no_ZI_LONGLAT_StARS" ))
```

# Running ZIPLN-network

### Fitting the model

#### Regional dataset

```{r regional_model_ZI}
regional_model<- ZIPLNnetwork(Abundance ~ 1 + PC1 + PC2,
                              data = regional_data_pc,
                              control = ZIPLNnetwork_param(min_ratio = 0.05,
                                                           penalize_diagonal = FALSE),
                               zi = "col")
```

```{r}
# regional_model_StARS <- regional_model$getBestModel("StARS")
regional_model_StARS <- regional_model$getModel(0.07299546)
```

#### Northern Range dataset

```{r northern_range_model_ZI}
northern_range_model <- ZIPLNnetwork(Abundance ~ 1 + PC1 + PC2,
                                     data = northern_range_data_pc,
                                     control = ZIPLNnetwork_param(min_ratio = 0.05,
                                                                 penalize_diagonal = FALSE),
                                     zi = "col")
```

```{r}
# northern_range_model_StARS <- northern_range_model$getBestModel("StARS")
northern_range_model_StARS <- northern_range_model$getModel(0.08051953)
```

### Goodness of fit

#### Regional dataset

```{r regional_fit_ZI}
Y_real   <- regional_data_pc$Abundance
Y_fitted <- regional_model_StARS$fitted ; colnames(Y_fitted) <- colnames(Y_real)
plot(Y_real, Y_fitted)
```

```{r}
print(rmse(Y_real, Y_fitted))
```

Focusing on the species with many 0s.

```{r regional_fit_zeros_ZI}
zero_threshold <- median(colSums(Y_real == 0))
zero_species   <- names(which(colSums(Y_real == 0) > zero_threshold & colSums(Y_real) < 400))
Y_zero         <- Y_real[, zero_species] ; Y_fitted_zero <- Y_fitted[, zero_species] 
plot(Y_zero, Y_fitted_zero)
```

```{r}
print(rmse(Y_zero, Y_fitted_zero))
```

#### Northern Range dataset

```{r nr_fit_ZI}
Y_real   <- northern_range_data$Abundance
Y_fitted <- northern_range_model_StARS$fitted ; colnames(Y_fitted) <- colnames(Y_real)
plot(Y_real, Y_fitted)
```

```{r}
print(rmse(Y_real, Y_fitted))
```

Focusing on the species with many 0s.

```{r nr_fit_zeros_ZI}
zero_threshold <- median(colSums(Y_real == 0))
zero_species   <- names(which(colSums(Y_real == 0) > zero_threshold & colSums(Y_real) < 400))
Y_zero         <- Y_real[, zero_species] ; Y_fitted_zero <- Y_fitted[, zero_species] 
plot(Y_zero, Y_fitted_zero)
```

```{r}
print(rmse(Y_zero, Y_fitted_zero))
```

### Resulting networks

#### Regional dataset

```{r}
plot(regional_model_StARS)
```

With a 0 penalty we would have:

```{r}
plot(regional_model$getModel(0))
```

Remark: we see that several associations are the same as those found with LONGLAT (Rhamquel - Coryriis, Poecreti - Anabhart, Sicypunc - Eleopiso).

#### Northern Range dataset

```{r}
plot(northern_range_model_StARS)
```

Remark: again, several associations are the same as those found with LONGLAT (Hoplmala - Anabhart, Rhamquel - Anabhart, Poecreti - Andipulc...).

With a 0 penalty we would have:

```{r}
plot(northern_range_model$getModel(0))
```

### Robustness of the networks to the addition of covariates

#### Regional network

```{r}
plot(regional_model_StARS)
```

```{r}
regional_PC3 <- ZIPLNnetwork(Abundance ~ 1 + PC1 + PC2 + PC3,
                             data = regional_data_pc,
                             control = ZIPLNnetwork_param(min_ratio = 0.05,
                                                            penalize_diagonal = FALSE),
                             zi = "col")
# regional_PC3_StARS <- regional_PC3$getBestModel("StARS")
regional_PC3_StARS <- regional_PC3$getModel(0.08177813)
# plot(regional_PC3_StARS)
```

```{r}
regional_LONGLAT <- ZIPLNnetwork(Abundance ~ 1 + LONGLAT,
                                 data = regional_data,
                                 control = ZIPLNnetwork_param(min_ratio = 0.05,
                                                            penalize_diagonal = FALSE),
                                 zi = "col")
# regional_LONGLAT_StARS <- regional_LONGLAT$getBestModel("StARS")
regional_LONGLAT_StARS <- regional_LONGLAT$getModel(0.06633124)
plot(regional_LONGLAT_StARS)
```

```{r}
regional_ziPC1 <- ZIPLNnetwork(Abundance ~ 1 + PC1 + PC2 | 1 + PC1,
                             data = regional_data_pc,
                             control = ZIPLNnetwork_param(min_ratio = 0.05,
                                                            penalize_diagonal = FALSE))
# regional_ziPC1_StARS <- regional_ziPC1$getBestModel("StARS")
regional_ziPC1_StARS <- regional_ziPC1$getModel(0.03697966)
plot(regional_ziPC1_StARS)
```

```{r}
regional_ziPC1PC2 <- ZIPLNnetwork(Abundance ~ 1 + PC1 + PC2 | 1 + PC1 + PC2,
                             data = regional_data_pc,
                             control = ZIPLNnetwork_param(min_ratio = 0.05,
                                                            penalize_diagonal = FALSE))
# regional_ziPC1PC2_StARS <- regional_ziPC1PC2$getBestModel("StARS")
regional_ziPC1PC2_StARS <- regional_ziPC1PC2$getModel(0.03828855)
# plot(regional_ziPC1_StARS)
```

```{r}
robustness(c(regional_model_StARS,
             regional_PC3_StARS,
             regional_LONGLAT_StARS,
             regional_ziPC1_StARS,
             regional_ziPC1PC2_StARS),
           c("regional_StARS",
             "regional_PC3_StARS",
             "regional_LONGLAT_StARS",
             "regional_ziPC1_StARS",
             "regional_ziPC1PC2_StARS" ))
```

#### Northern Range network

```{r}
plot(northern_range_model_StARS)
```

```{r}
# northern_range_PC3 <- ZIPLNnetwork(Abundance ~ 1 + PC1 + PC2 + PC3,
#                              data = northern_range_data_pc,
#                              control = ZIPLNnetwork_param(min_ratio = 0.05,
#                                                             penalize_diagonal = FALSE),
#                              zi = "col")
# regional_PC3_StARS <- regional_PC3$getBestModel("StARS")
# regional_PC3_StARS <- regional_PC3$getModel(0.08177813)
# plot(regional_PC3_StARS)
```

```{r}
northern_range_LONGLAT <- ZIPLNnetwork(Abundance ~ 1 + LONGLAT,
                                               data = northern_range_data,
                                               control = ZIPLNnetwork_param(min_ratio = 0.05,
                                                            penalize_diagonal = FALSE),
                                               zi = "col")
# northern_range_LONGLAT_StARS <- northern_range_LONGLAT$getBestModel("StARS")
northern_range_LONGLAT_StARS <- northern_range_LONGLAT$getModel(0.08978182)
# plot(regional_LONGLAT_StARS)
```

```{r}
northern_range_ziPC1 <- ZIPLNnetwork(Abundance ~ 1 + PC1 + PC2 | 1 + PC1,
                                     data = northern_range_data_pc,
                                     control = ZIPLNnetwork_param(min_ratio = 0.05,
                                                            penalize_diagonal = FALSE))
# northern_range_ziPC1_StARS <- northern_range_ziPC1$getBestModel("StARS")
northern_range_ziPC1_StARS <- northern_range_ziPC1$getModel(0.04591576)
plot(northern_range_ziPC1_StARS)
```

```{r}
# Error in if (parameters_converged | objective_converged)
# northern_range_ziPC1PC2 <- ZIPLNnetwork(Abundance ~ 1 + PC1 + PC2 | 1 + PC1 + PC2,
#                              data = northern_range_data_pc,
#                              control = ZIPLNnetwork_param(min_ratio = 0.05,
#                                                             penalize_diagonal = FALSE))
# northern_range_ziPC1PC2_StARS <- northern_range_ziPC1PC2$getBestModel("StARS")
# northern_range_ziPC1PC2_StARS   <- northern_range_ziPC1PC2$getModel(0.03828855)
# plot(northern_range_ziPC1PC2_StARS)
```

```{r}
robustness(c(northern_range_model_StARS,
             northern_range_LONGLAT_StARS,
             northern_range_ziPC1_StARS),
           c("northern_range_StARS",
             "northern_range_LONGLAT_StARS",
             "northern_range_ziPC1_StARS"))
```

# Accounting for sampling date in the Northern Range dataset

## Formatting date as appropriate covariates

```{r}
date_covar <- data.frame("Year"  = as.factor(NORTHERN_RANGE_DATASET_COVARIATES$YEAR),
                         "Month" = as.factor(NORTHERN_RANGE_DATASET_COVARIATES$MONTH))
northern_range_date_pc_covar  <- cbind(date_covar, pca_covar_northern_range)
rownames(northern_range_date_pc_covar) <- rownames(NORTHERN_RANGE_DATASET_COVARIATES)
northern_range_data_date_pc <- prepare_data(counts     = NORTHERN_RANGE_DATASET_COUNTS,
                                            covariates = northern_range_date_pc_covar)
```

## Re-running PLN-network with the date covariates

```{r}
northern_range_no_ZI_model_year <- PLNnetwork(Abundance ~ 1 + PC1 + PC2 + Year,
                                              data = northern_range_data_date_pc,
                                              control = PLNnetwork_param(min_ratio = 0.05,
                                                           penalize_diagonal = FALSE))
# northern_range_no_ZI_model_year_StARS <- northern_range_no_ZI_model_year$getBestModel("StARS")
northern_range_no_ZI_model_year_StARS <- northern_range_no_ZI_model_year$getModel(0.5421913)
# plot(northern_range_no_ZI_model_year_StARS)
```

```{r}
northern_range_no_ZI_model_year_month <- PLNnetwork(Abundance ~ 1 + PC1 + PC2 + Year +  Month,
                                                    data = northern_range_data_date_pc,
                                                    control = PLNnetwork_param(min_ratio = 0.05,
                                                              penalize_diagonal = FALSE))
# northern_range_no_ZI_model_year_month_StARS <- northern_range_no_ZI_model_year_month$getBestModel("StARS")
northern_range_no_ZI_model_year_month_StARS <- northern_range_no_ZI_model_year_month$getModel(0.3764233)
# plot(northern_range_model_year_month_StARS)
```

## Re-running ZIPLN-network with the date covariates

```{r}
northern_range_model_year <- ZIPLNnetwork(Abundance ~ 1 + PC1 + PC2 + Year,
                                          data = northern_range_data_date_pc,
                                          control = ZIPLNnetwork_param(min_ratio = 0.05,
                                                           penalize_diagonal = FALSE),
                                          zi = "col")
# northern_range_model_year_StARS <- northern_range_model_year$getBestModel("StARS")
northern_range_model_year_StARS <- northern_range_model_year$getModel(0.07216102)
plot(northern_range_model_year_StARS)
```

Remark: at first sight, it appears that adding the Year as a covariate does not significantly change the network structure.
```{r}
northern_range_model_year_ZIyear <- ZIPLNnetwork(Abundance ~ 1 + PC1 + PC2 + Year | 1 + Year,
                                          data = northern_range_data_date_pc,
                                          control = ZIPLNnetwork_param(min_ratio = 0.05,
                                                           penalize_diagonal = FALSE))
# northern_range_model_year_ZIyear_StARS <- northern_range_model_year_ZIyear$getBestModel("StARS")
northern_range_model_year_ZIyear_StARS <- northern_range_model_year_ZIyear$getModel(0.07185885)
plot(northern_range_model_year_ZIyear_StARS)
```

```{r}
# northern_range_model_month <- ZIPLNnetwork(Abundance ~ 1 + PC1 + PC2 + Month,
#                                           data = northern_range_data_date_pc,
#                                           control = ZIPLNnetwork_param(min_ratio = 0.05,
#                                                            penalize_diagonal = FALSE),
#                                           zi = "col")
# Error in if (parameters_converged | objective_converged) { : 
# missing value where TRUE/FALSE needed
```

```{r}
northern_range_model_year_month <- ZIPLNnetwork(Abundance ~ 1 + PC1 + PC2 + Year +  Month,
                                               data = northern_range_data_date_pc,
                                               control = ZIPLNnetwork_param(min_ratio = 0.05,
                                                           penalize_diagonal = FALSE),
                                               zi = "col")
# northern_range_model_year_month_StARS <- northern_range_model_year_month$getBestModel("StARS")
northern_range_model_year_month_StARS <- northern_range_model_year_month$getModel(0.05944257)
plot(northern_range_model_year_month_StARS)
```

Remark: one significant change is the disappearing of the Rhamquel - Anabhart association. A few associations lose strength (Hoplamala - Rhamquel) while others seem stronger (Ancimara - Hyporobi).

```{r}
northern_range_model_ZImonth <- ZIPLNnetwork(Abundance ~ 1 + PC1 + PC2 | 1 +  Month,
                                          data = northern_range_data_date_pc,
                                          control = ZIPLNnetwork_param(min_ratio = 0.05,
                                                           penalize_diagonal = FALSE))
# northern_range_model_ZImonth_StARS <- northern_range_model_ZImonth$getBestModel("StARS")
northern_range_model_ZImonth_StARS <- northern_range_model_ZImonth$getModel(0.08764195)
plot(northern_range_model_ZImonth_StARS)
```

# Comparison of the networks with / without ZI

```{r}
robustness(c(regional_no_ZI_StARS,
             regional_no_ZI_PC3_StARS,
             regional_no_ZI_LONGLAT_StARS,
             regional_model_StARS,
             regional_PC3_StARS,
             regional_LONGLAT_StARS,
             regional_ziPC1_StARS,
             regional_ziPC1PC2_StARS),
           c("regional_no_ZI_StARS",
             "regional_no_ZI_PC3_StARS",
             "regional_no_ZI_LONGLAT_StARS",
             "regional_StARS",
             "regional_PC3_StARS",
             "regional_LONGLAT_StARS",
             "regional_ziPC1_StARS",
             "regional_ziPC1PC2_StARS" ))
```

```{r}
robustness(c(northern_range_no_ZI_StARS,
             northern_range_no_ZI_LONGLAT_StARS,
             northern_range_no_ZI_model_year_StARS,
             northern_range_no_ZI_model_year_month_StARS,
             northern_range_model_StARS,
             northern_range_LONGLAT_StARS,
             northern_range_ziPC1_StARS,
             northern_range_model_year_StARS,
             northern_range_model_year_month_StARS),
           c("northern_range_no_ZI_StARS",
             "northern_range_no_ZI_LONGLAT_StARS",
             "northern_range_no_ZI_model_year_StARS",
             "northern_range_no_ZI_model_year_month_StARS",
             "northern_range_StARS",
             "northern_range_LONGLAT_StARS",
             "northern_range_ziPC1_StARS",
             "northern_range_model_year_StARS",
             "northern_range_model_year_month_StARS"))
```

# Comparison with Hmsc package
```{r}
Y_regional <- regional_data$Abundance
Y_NR       <- northern_range_data$Abundance
```

## With LONGLAT
### Regional data
```{r}
X_regional_LONGLAT    <- REGIONAL_DATASET_COVARIATES %>%  select(c("LONGLAT"))

nChains = 2
thin = 10
samples = 1000
transient = 500*thin
verbose = 0
n = nrow(Y_regional)

studyDesign = data.frame(sample = as.factor(1:n))
rL = HmscRandomLevel(units = studyDesign$sample)
m_regional_LONGLAT = Hmsc(Y = Y_regional, XData = X_regional_LONGLAT, XFormula = ~LONGLAT,
studyDesign = studyDesign, ranLevels = list(sample = rL), distr = "lognormal poisson")
m_regional_LONGLAT  = sampleMcmc(m_regional_LONGLAT , thin = thin, samples = samples, transient = transient,
nChains = nChains, nParallel = nChains, verbose = verbose)
```

```{r}
preds = computePredictedValues(m_regional_LONGLAT )
evaluateModelFit(hM = m_regional_LONGLAT , predY = preds)
```

```{r}
OmegaCor = computeAssociations(m_regional_LONGLAT)
supportLevel = 0.95
Sigma_regional_LONGLAT = ((OmegaCor[[1]]$support>supportLevel)
+ (OmegaCor[[1]]$support<(1-supportLevel))>0)*OmegaCor[[1]]$mean
corrplot(Sigma_regional_LONGLAT, method = "color",
col = colorRampPalette(c("blue","white","red"))(200),
title = paste("random effect level:", m_regional_LONGLAT $rLNames[1]), mar=c(0,0,1,0))
```
```{r}
plot_network(Sigma_regional_LONGLAT * 0.1)
```

### Northern Range data
```{r}
X_NR_LONGLAT    <- northern_range_data %>%  select(c("LONGLAT"))
nChains = 2
thin = 10
samples = 1000
transient = 500*thin
verbose = 0
n = nrow(Y_NR)

studyDesign = data.frame(sample = as.factor(1:n))
rL = HmscRandomLevel(units = studyDesign$sample)
m_NR_LONGLAT = Hmsc(Y = Y_NR, XData = X_NR_LONGLAT, XFormula = ~LONGLAT,
studyDesign = studyDesign, ranLevels = list(sample = rL), distr = "lognormal poisson")
m_NR_LONGLAT = sampleMcmc(m_NR_LONGLAT, thin = thin, samples = samples, transient = transient,
nChains = nChains, nParallel = nChains, verbose = verbose)
```

```{r}
preds = computePredictedValues(m_NR_LONGLAT)
evaluateModelFit(hM = m_NR_LONGLAT, predY = preds)
```

```{r}
OmegaCor = computeAssociations(m_NR_LONGLAT)
supportLevel = 0.95
Sigma_NR_LONGLAT = ((OmegaCor[[1]]$support>supportLevel)
+ (OmegaCor[[1]]$support<(1-supportLevel))>0)*OmegaCor[[1]]$mean
corrplot(Sigma_NR_LONGLAT, method = "color",
col = colorRampPalette(c("blue","white","red"))(200),
title = paste("random effect level:", m_NR_LONGLAT$rLNames[1]), mar=c(0,0,1,0))
```

```{r}
plot_network(Sigma_NR_LONGLAT*0.1)
```

## With PC1 + PC1
### Regional data
```{r}
X_regional_PC1_PC2    <- regional_data_pc %>%  select(c("PC1", "PC2"))

nChains = 2
thin = 10
samples = 1000
transient = 500*thin
verbose = 0
n = nrow(Y_regional)

studyDesign = data.frame(sample = as.factor(1:n))
rL = HmscRandomLevel(units = studyDesign$sample)
m_regional_PC1_PC2 = Hmsc(Y = Y_regional, XData = X_regional_PC1_PC2, XFormula = ~ PC1 + PC2,
studyDesign = studyDesign, ranLevels = list(sample = rL), distr = "lognormal poisson")
m_regional_PC1_PC2  = sampleMcmc(m_regional_PC1_PC2, thin = thin, samples = samples, transient = transient,
nChains = nChains, nParallel = nChains, verbose = verbose)
```

```{r}
preds = computePredictedValues(m_regional_PC1_PC2 )
evaluateModelFit(hM = m_regional_PC1_PC2 , predY = preds)
```

```{r}
OmegaCor = computeAssociations(m_regional_PC1_PC2)
supportLevel = 0.95
Sigma_regional_PC1_PC2 = ((OmegaCor[[1]]$support>supportLevel)
+ (OmegaCor[[1]]$support<(1-supportLevel))>0)*OmegaCor[[1]]$mean
corrplot(Sigma_regional_PC1_PC2, method = "color",
col = colorRampPalette(c("blue","white","red"))(200),
title = paste("random effect level:", m_regional_LONGLAT $rLNames[1]), mar=c(0,0,1,0))
```

```{r}
plot_network(Sigma_regional_PC1_PC2 * 0.1)
```

### Northern Range data
```{r}
X_NR_PC1_PC2    <- northern_range_data_pc %>%  select(c("PC1", "PC2"))

nChains = 2
thin = 10
samples = 1000
transient = 500*thin
verbose = 0
n = nrow(Y_NR)

studyDesign = data.frame(sample = as.factor(1:n))
rL = HmscRandomLevel(units = studyDesign$sample)
m_NR_PC1_PC2 = Hmsc(Y = Y_NR, XData = X_NR_PC1_PC2, XFormula = ~PC1 + PC2,
studyDesign = studyDesign, ranLevels = list(sample = rL), distr = "lognormal poisson")
m_NR_PC1_PC2 = sampleMcmc(m_NR_PC1_PC2, thin = thin, samples = samples, transient = transient,
nChains = nChains, nParallel = nChains, verbose = verbose)
```

```{r}
preds = computePredictedValues(m_NR_PC1_PC2)
evaluateModelFit(hM = m_NR_PC1_PC2, predY = preds)
```

```{r}
OmegaCor = computeAssociations(m_NR_PC1_PC2)
supportLevel = 0.95
Sigma_NR_PC1_PC2 = ((OmegaCor[[1]]$support>supportLevel)
+ (OmegaCor[[1]]$support<(1-supportLevel))>0)*OmegaCor[[1]]$mean
corrplot(Sigma_NR_PC1_PC2, method = "color",
col = colorRampPalette(c("blue","white","red"))(200),
title = paste("random effect level:", m_NR_PC1_PC2$rLNames[1]), mar=c(0,0,1,0))
```

```{r}
plot_network(Sigma_NR_PC1_PC2 * 0.1)
```

## With PC1 + PC2 + dates
### Just year
```{r}
X_NR_PC1_PC2_year    <- northern_range_data_date_pc %>%  select(c("PC1", "PC2", "Year"))

nChains = 2
thin = 10
samples = 1000
transient = 500*thin
verbose = 0
n = nrow(Y_NR)

studyDesign = data.frame(sample = as.factor(1:n))
rL = HmscRandomLevel(units = studyDesign$sample)
m_NR_PC1_PC2_year = Hmsc(Y = Y_NR, XData = X_NR_PC1_PC2_year, XFormula = ~PC1 + PC2 + Year,
studyDesign = studyDesign, ranLevels = list(sample = rL), distr = "lognormal poisson")
m_NR_PC1_PC2_year = sampleMcmc(m_NR_PC1_PC2_year, thin = thin, samples = samples, transient = transient,
nChains = nChains, nParallel = nChains, verbose = verbose)
```

```{r}
preds = computePredictedValues(m_NR_PC1_PC2_year)
evaluateModelFit(hM = m_NR_PC1_PC2_year, predY = preds)
```

```{r}
OmegaCor = computeAssociations(m_NR_PC1_PC2_year)
supportLevel = 0.95
Sigma_NR_PC1_PC2_Year = ((OmegaCor[[1]]$support>supportLevel)
+ (OmegaCor[[1]]$support<(1-supportLevel))>0)*OmegaCor[[1]]$mean
corrplot(Sigma_NR_PC1_PC2_Year, method = "color",
col = colorRampPalette(c("blue","white","red"))(200),
title = paste("random effect level:", m_NR_PC1_PC2_year$rLNames[1]), mar=c(0,0,1,0))
```

```{r}
plot_network(Sigma_NR_PC1_PC2_Year*0.1)
```

### Just month
```{r}
X_NR_PC1_PC2_month     <- northern_range_data_date_pc %>%  select(c("PC1", "PC2", "Month"))

nChains = 2
thin = 10
samples = 1000
transient = 500*thin
verbose = 0
n = nrow(Y_NR)

studyDesign = data.frame(sample = as.factor(1:n))
rL = HmscRandomLevel(units = studyDesign$sample)
m_NR_PC1_PC2_month = Hmsc(Y = Y_NR, XData = X_NR_PC1_PC2_month, XFormula = ~ PC1 + PC2 + Month,
studyDesign = studyDesign, ranLevels = list(sample = rL), distr = "lognormal poisson")
m_NR_PC1_PC2_month = sampleMcmc(m_NR_PC1_PC2_month, thin = thin, samples = samples, transient = transient,
nChains = nChains, nParallel = nChains, verbose = verbose)
```

```{r}
preds = computePredictedValues(m_NR_PC1_PC2_month)
evaluateModelFit(hM = m_NR_PC1_PC2_month, predY = preds)
```

```{r}
OmegaCor = computeAssociations(m_NR_PC1_PC2_month)
supportLevel = 0.95
Sigma_NR_PC1_PC2_Month = ((OmegaCor[[1]]$support>supportLevel)
+ (OmegaCor[[1]]$support<(1-supportLevel))>0)*OmegaCor[[1]]$mean
corrplot(Sigma_NR_PC1_PC2_Month, method = "color",
col = colorRampPalette(c("blue","white","red"))(200),
title = paste("random effect level:", m_NR_PC1_PC2_month$rLNames[1]), mar=c(0,0,1,0))
```

```{r}
# Omega <- solve(Sigma_NR_PC1_PC2_Month)
# partial_corr <- - Omega / tcrossprod(sqrt(diag(Omega))); diag(partial_corr) <- 1
# plot_network(partial_corr)
plot_network(Sigma_NR_PC1_PC2_Month * 0.1)
```

### Year + Month
```{r}
X_NR_PC1_PC2_year_month     <- northern_range_data_date_pc %>%  select(c("PC1", "PC2", "Year", "Month"))
nChains = 2
thin = 10
samples = 1000
transient = 500*thin
verbose = 0
n = nrow(Y_NR)

studyDesign = data.frame(sample = as.factor(1:n))
rL = HmscRandomLevel(units = studyDesign$sample)
m_NR_PC1_PC2_year_month = Hmsc(Y = Y_NR, XData = X_NR_PC1_PC2_year_month, XFormula = ~ PC1 + PC2 + Year + Month,
studyDesign = studyDesign, ranLevels = list(sample = rL), distr = "lognormal poisson")
m_NR_PC1_PC2_year_month = sampleMcmc(m_NR_PC1_PC2_year_month, thin = thin, samples = samples, transient = transient,
nChains = nChains, nParallel = nChains, verbose = verbose)
```

```{r}
preds = computePredictedValues(m_NR_PC1_PC2_year_month)
evaluateModelFit(hM = m_NR_PC1_PC2_year_month, predY = preds)
```

```{r}
OmegaCor = computeAssociations(m_NR_PC1_PC2_year_month)
supportLevel = 0.95
Sigma_NR_PC1_PC2_Year_Month = ((OmegaCor[[1]]$support>supportLevel)
+ (OmegaCor[[1]]$support<(1-supportLevel))>0)*OmegaCor[[1]]$mean
corrplot(Sigma_NR_PC1_PC2_Year_Month, method = "color",
col = colorRampPalette(c("blue","white","red"))(200),
title = paste("random effect level:", m_NR_PC1_PC2_year_month$rLNames[1]), mar=c(0,0,1,0))
```

```{r}
plot_network(Sigma_NR_PC1_PC2_Year_Month*0.1)
```

## Comparing the HMSC networks together
### Regional network
```{r}
res <- robustness(list(Sigma_regional_LONGLAT,
                    Sigma_regional_PC1_PC2),
                  c("Sigma_regional_LONGLAT",
                    "Sigma_regional_PC1_PC2"),
                  matrices_input = TRUE
                  )
print(res)
```

### Northern Range network
```{r}
res <- robustness(list(Sigma_NR_LONGLAT,
                    Sigma_NR_PC1_PC2,
                    Sigma_NR_PC1_PC2_Year,
                    Sigma_NR_PC1_PC2_Month,
                    Sigma_NR_PC1_PC2_Year_Month),
                  c("Sigma_NR_LONGLAT",
                    "Sigma_NR_PC1_PC2",
                    "Sigma_NR_PC1_PC2_Year",
                    "Sigma_NR_PC1_PC2_Month",
                    "Sigma_NR_PC1_PC2_Year_Month"),
                  matrices_input = TRUE
                  )
print(res)
```

## Comparing the HMSC networks with the PLN ones
### Regional network
```{r}
res <- robustness(list(Sigma_regional_LONGLAT,
                       Sigma_regional_PC1_PC2,
                       as.matrix(regional_no_ZI_StARS$latent_network("partial_cor")),
                       as.matrix(regional_no_ZI_PC3_StARS$latent_network("partial_cor")),
                       as.matrix(regional_no_ZI_LONGLAT_StARS$latent_network("partial_cor")),
                       as.matrix(regional_model_StARS$latent_network("partial_cor")),
                       as.matrix(regional_PC3_StARS$latent_network("partial_cor")),
                       as.matrix(regional_LONGLAT_StARS$latent_network("partial_cor")),
                       as.matrix(regional_ziPC1_StARS$latent_network("partial_cor")),
                       as.matrix(regional_ziPC1PC2_StARS$latent_network("partial_cor"))),
                  c("HMSC_Sigma_regional_LONGLAT",
                    "HMSC_Sigma_regional_PC1_PC2",
                    "regional_no_ZI_StARS",
                    "regional_no_ZI_PC3_StARS",
                    "regional_no_ZI_LONGLAT_StARS",
                    "regional_model_StARS",
                    "regional_PC3_StARS",
                    "regional_LONGLAT_StARS",
                    "regional_ziPC1_StARS",
                    "regional_ziPC1PC2_StARS"),
                  matrices_input = TRUE
                  )
print(res)
```

### Northern Range network
```{r}
res <- robustness(list(Sigma_NR_LONGLAT,
                       Sigma_NR_PC1_PC2,
                       Sigma_NR_PC1_PC2_Year,
                       Sigma_NR_PC1_PC2_Month,
                       Sigma_NR_PC1_PC2_Year_Month,
                       as.matrix(northern_range_no_ZI_StARS$latent_network("partial_cor")),
                       as.matrix(northern_range_no_ZI_LONGLAT_StARS$latent_network("partial_cor")),
                       as.matrix(northern_range_no_ZI_model_year_StARS$latent_network("partial_cor")),
                       as.matrix(northern_range_no_ZI_model_year_month_StARS$latent_network("partial_cor")),
                       as.matrix(northern_range_model_StARS$latent_network("partial_cor")),
                       as.matrix(northern_range_LONGLAT_StARS$latent_network("partial_cor")),
                       as.matrix(northern_range_ziPC1_StARS$latent_network("partial_cor")),
                       as.matrix(northern_range_model_year_StARS$latent_network("partial_cor")),
                       as.matrix(northern_range_model_year_month_StARS$latent_network("partial_cor"))),
                  c("HSMC_Sigma_NR_LONGLAT",
                    "HSMC_Sigma_NR_PC1_PC2",
                    "HSMC_Sigma_NR_PC1_PC2_Year",
                    "HSMC_Sigma_NR_PC1_PC2_Month",
                    "HSMC_Sigma_NR_PC1_PC2_Year_Month",
                    "northern_range_no_ZI_StARS",
                    "northern_range_no_ZI_LONGLAT_StARS",
                    "northern_range_no_ZI_model_year_StARS",
                    "northern_range_no_ZI_model_year_month_StARS",
                    "northern_range_model_StARS",
                    "northern_range_LONGLAT_StARS",
                    "northern_range_ziPC1_StARS",
                    "northern_range_model_year_StARS",
                    "northern_range_model_year_month_StARS"),
                  matrices_input = TRUE
                  )

print(res)
```

### 2025-06-10 DEPRECATED/PAPER_CODE_files/PAPER_CODE.qmd
rewrote the code to adjust for the changes in the v3 draft

--- 2025-03-21 PAPER_CODE.qmd
@@
 library(dplyr)
-library(PLNmodels)
+# library(PLNmodels)
 library(igraph)
@@
   }
-  plot(G, layout = layout)
+  plot(G, layout = layout, font.size = 8)
   invisible(G)

### 2025-06-10 PAPER_CODE.qmd
rewrote the code to adjust for the changes in the v3 draft

--- 2025-03-19 PAPER_CODE.qmd
@@
 
-## Useful libraries & functions
+# Useful libraries & functions
+
+PLNmodels version should be 1.2.2 or higher.
+
+```{r, loading libraries}
+# library(PLNmodels)
+library(Metrics)
+library(pheatmap)
+library(tidyverse)
+library(FactoMineR)
+library(factoextra)
+library(viridisLite)
+```
+
+# Loading & formatting data
+
+## Data loading
+
+```{r data-loading}
+load("data/NORTHERN_RANGE_DATASET_COUNTS.Rdata")
+NR_COUNTS <- NORTHERN_RANGE_DATASET_COUNTS %>% as_tibble()
+NR_COUNTS <- NR_COUNTS[, -21]
+
+load("data/NORTHERN_RANGE_DATASET_COVARIATES.Rdata")
+NR_COVARIATES <- 
+  NORTHERN_RANGE_DATASET_COVARIATES %>% as_tibble() %>% 
+  rename(latitude = LATITUDE, 
+         longitude = LONGITUDE, 
+         long_lat = LONGLAT,
+         coarse_gravel = coarse.gravel, # We do not include geographical coordinates here
+         fine_gravel = fine.gravel, 
+         leaf_litter = leaf.litter,
+         time_step = TimeStep, year = YEAR, month = MONTH, 
+         stream = STREAM, 
+         disturbance = DISTURBANCE, # human activity / binary categorical variable
+         altitude = ALTITUDE, 
+         season = SEASON) %>% 
+  mutate(
+    # season = case_match(time_step,
+    #   c(1, 5,  9, 13, 17) ~ "dry_start",
+    #   c(2, 6, 10, 14, 18) ~ "dry_end",
+    #   c(3, 7, 11, 15, 19) ~ "wet_start",
+    #   c(4, 8, 12, 16    ) ~ "wet_end"
+    # ),
+    season = as.character(season),
+    year = as.character(year),
+    disturbance = ifelse(disturbance == 1, "yes", "no")
+  ) %>% 
+  ## flow is not reliable
+  ## time_step + month are redundant with season + year
+  ## site is redudant with disturbance + stream
+  ## long_lat is the 1st PCA axis of latitude + longitude
+  dplyr::select(-latitude, -longitude, - long_lat, -site, -time_step,
+                -month, -flow) %>% # we remove geographical coordinates
+  relocate(altitude, width, depth, volume, garbage,
+           conductivity, O2, pH, temperature, turbidity, # site features
+           coarse_gravel, fine_gravel, leaf_litter, cobble, sand, silt, boulders, canopy, # soil
+           season, year, # sampling time
+           stream, disturbance # categorical: place + human activity
+         )
+# turbidity: ordinal variable indicating the level of turbidity
+# season:-> dry-start/dry-end, wet-start/wet-end, (january-> may: dry)
+quali_ind <- seq(ncol(NR_COVARIATES) - 3, ncol(NR_COVARIATES))
+
+```
+
+## Data visualization
 
 ```{r}
-library(dplyr)
-# library(PLNmodels)
-library(igraph)
+log_abundance <- log(1 + NR_DATA$Abundance)
+log_abundance_sorted <- log_abundance[order(NR_DATA$stream), ]
+
+color_palette <- paletteer_c("ggthemes::Orange-Gold", n = 30)
+palette_colors <- paletteer::paletteer_c("ggthemes::Orange-Gold", n = 29)  # 29 instead of 30
+color_palette <- c("white", palette_colors)
+max_val <- max(log_abundance_sorted, na.rm = TRUE)
+
+pheatmap(log_abundance_sorted,
+         color = color_palette,
+         cluster_rows = FALSE, cluster_cols = FALSE,
+         annotation_row = data.frame(Stream = NR_DATA$stream),
+         show_rownames = FALSE)
+```
+## Covariates preparation (PCA)
+
+```{r pca}
+nb_pc <- ncol(NR_COVARIATES) - 4
+NR_PCA <-
+  NR_COVARIATES %>%
+  PCA(quali.sup = quali_ind, ncp = nb_pc, scale.unit = TRUE, graph = FALSE)
+
+NR_PCA_SCORES <- setNames(data.frame(NR_PCA$ind$coord), paste0("PC",1:nb_pc))
+```
+
+```{r data-formatting}
+NR_DATA <- prepare_data(
+  counts     = NR_COUNTS,
+  covariates = cbind(NR_COVARIATES, NR_PCA_SCORES)
+  )
+```
+
+# Running PLN-network
+
+```{r pln_network_pc1_pc2}
+model_PLN_pc1_pc2  <- PLNnetwork(Abundance ~ PC1 + PC2,  data = NR_DATA,
+                         control = PLNnetwork_param(min_ratio = 0.05,
+                                                    penalize_diagonal = FALSE))
+
+# model_PLN_pc1_pc2_StARS <- model_PLN_pc1_pc1$getBestModel("StARS", stability = 0.8)
+model_PLN_pc1_pc2_StARS <- model_PLN_pc1_pc2$getModel(0.2444352)
+print(model_PLN_pc1_pc2_StARS$BIC)
+
+model_PLN_pc1_pc2_StARS$plot_network()
+
+B_log_scale <- sign(model_PLN_pc1_pc2_StARS$model_par$B) * log1p(abs(model_PLN_pc1_pc2_StARS$model_par$B))
+rownames(B_log_scale) <- c("Intercept", "PC1", "PC2")
+
+pheatmap(B_log_scale,color = paletteer::paletteer_c("viridis::viridis", 30),
+         cluster_rows = FALSE, cluster_cols = FALSE)
+```
+
+```{r pln_network_pc1_pc1_disturbance}
+model_PLN_pc1_pc2_disturbance  <- PLNnetwork(Abundance ~ PC1 + PC2 + disturbance,  data = NR_DATA,
+                         control = PLNnetwork_param(min_ratio = 0.05,
+                                                    penalize_diagonal = FALSE))
+# model_PLN_pc1_pc2_disturbance_StARS <- model_PLN_pc1_pc2_disturbance$getBestModel("StARS", stability = 0.8)
+model_PLN_pc1_pc2_disturbance_StARS <- model_PLN_pc1_pc2_disturbance$getModel(0.2361045)
+print(model_PLN_pc1_pc2_disturbance_StARS$BIC)
+
+model_PLN_pc1_pc2_disturbance_StARS$plot_network()
+
+B_log_scale <- sign(model_PLN_pc1_pc2_disturbance_StARS$model_par$B) * log1p(abs(model_PLN_pc1_pc2_disturbance_StARS$model_par$B))
+rownames(B_log_scale) <- c("Intercept", "PC1", "PC2", "disturbance")
+pheatmap(B_log_scale,
+         color = paletteer::paletteer_c("viridis::viridis", 30),
+         cluster_rows = FALSE, cluster_cols = FALSE)
+```
+
+```{r pln_network}
+model_PLN  <- PLNnetwork(Abundance ~ PC1 + PC2 + stream + disturbance,  data = NR_DATA,
+                         control = PLNnetwork_param(min_ratio = 0.05,
+                                                    penalize_diagonal = FALSE))
+# model_PLN_StARS <- model_PLN$getBestModel("StARS", stability = 0.8)
+model_PLN_StARS <- model_PLN$getModel(0.06570186)
+print(model_PLN_StARS$BIC)
+
+model_PLN_StARS$plot_network()
+
+B_log_scale <- sign(model_PLN_StARS$model_par$B) * log1p(abs(model_PLN_StARS$model_par$B))
+rownames(B_log_scale) <- c("Intercept", "PC1", "PC2", "stream_Caura", "stream_Lopinot",
+                           "stream_LowerAripo", "stream_Maracas",   "stream_Quare",
+                           "stream_Turure", "stream_UpperAripo", "disturbance")
+pheatmap(B_log_scale,
+         color = paletteer::paletteer_c("viridis::viridis", 30),
+         cluster_rows = FALSE, cluster_cols = FALSE)
+```
+
+# Running ZIPLN-network
+
+```{r zipln_network_pc1_pc2 }
+model_ZIPLN_pc1_pc2   <- ZIPLNnetwork(Abundance ~ PC1 + PC2 | 1,  data = NR_DATA,
+                         control = ZIPLNnetwork_param(min_ratio = 0.05,
+                                                      penalize_diagonal = FALSE))
+# model_ZIPLN_pc1_pc2_StARS <- model_ZIPLN_pc1_pc2$getBestModel("StARS", stability = 0.8)
+model_ZIPLN_pc1_pc2_StARS <- model_ZIPLN_pc1_pc2$getModel(0.05578001)
+print(model_ZIPLN_pc1_pc2_StARS$BIC)
+
+model_ZIPLN_pc1_pc2_StARS$plot_network()
+
+B_log_scale <- sign(model_ZIPLN_pc1_pc2_StARS$model_par$B) * log1p(abs(model_ZIPLN_pc1_pc2_StARS$model_par$B))
+rownames(B_log_scale) <- c("Intercept", "PC1", "PC2")
+pheatmap(B_log_scale,
+         color = paletteer::paletteer_c("viridis::viridis", 30),
+         cluster_rows = FALSE, cluster_cols = FALSE)
+
+pheatmap(model_ZIPLN_pc1_pc2_StARS$model_par$Pi,
+          color = paletteer::paletteer_c("viridis::cividis", 30),
+          cluster_rows = FALSE, cluster_cols = FALSE)
+```
+
+```{r zipln_network_pc1_pc2_disturbance }
+model_ZIPLN_pc1_pc2_disturbance   <- ZIPLNnetwork(Abundance ~ PC1 + PC2 + disturbance | 1,  data = NR_DATA,
+                                                  control = ZIPLNnetwork_param(min_ratio = 0.05,
+                                                      penalize_diagonal = FALSE))
+# model_ZIPLN_pc1_pc2_disturbance_StARS <- model_ZIPLN_pc1_pc2_disturbance$getBestModel("StARS", stability = 0.8)
+model_ZIPLN_pc1_pc2_disturbance_StARS <- model_ZIPLN_pc1_pc2_disturbance$getModel(0.04622685)
+print(model_ZIPLN_pc1_pc2_disturbance_StARS$BIC)
+
+model_ZIPLN_pc1_pc2_disturbance_StARS$plot_network()
+
+B_log_scale <- sign(model_ZIPLN_pc1_pc2_disturbance_StARS$model_par$B) * log1p(abs(model_ZIPLN_pc1_pc2_disturbance_StARS$model_par$B))
+rownames(B_log_scale) <- c("Intercept", "PC1", "PC2", "disturbance")
+pheatmap(B_log_scale,
+         color = paletteer::paletteer_c("viridis::viridis", 30),
+         cluster_rows = FALSE, cluster_cols = FALSE)
+
+pheatmap(model_ZIPLN_pc1_pc2_disturbance_StARS$model_par$Pi,
+         color = paletteer::paletteer_c("viridis::cividis", 30),
+         cluster_rows = FALSE, cluster_cols = FALSE)
+```
+
+```{r zipln_network}
+model_ZIPLN  <- ZIPLNnetwork(Abundance ~ PC1 + PC2 + disturbance | stream,  data = NR_DATA,
+                         control = ZIPLNnetwork_param(min_ratio = 0.05,
+                                                      penalize_diagonal = FALSE))
+# model_ZIPLN_StARS <- model_ZIPLN$getBestModel("StARS", stability = 0.8)
+model_ZIPLN_StARS <- model_ZIPLN$getModel(0.03317969)
+print(model_ZIPLN_StARS$BIC)
+
+model_ZIPLN_StARS$plot_network()
+
+B_log_scale <- sign(model_ZIPLN_StARS$model_par$B) * log1p(abs(model_ZIPLN_StARS$model_par$B))
+rownames(B_log_scale) <- c("Intercept", "PC1", "PC2", "disturbance")
+pheatmap(B_log_scale,
+         color = paletteer::paletteer_c("viridis::viridis", 30),
+         cluster_rows = FALSE, cluster_cols = FALSE)
+
+pheatmap(model_ZIPLN_pc1_pc2_disturbance_StARS$model_par$Pi,
+         color = paletteer::paletteer_c("viridis::cividis", 30),
+         cluster_rows = FALSE, cluster_cols = FALSE)
 ```
@@
 ```{r}
-plot_network = function(net             ,
-                         type            ,
-                         output          ,
-                         edge.color      = c("#F8766D", "#00BFC4"),
-                         remove.isolated = FALSE,
-                         node.labels     = NULL,
-                         layout          = layout_in_circle,
-                         plot = TRUE) {
+pi_proba <- model_ZIPLN_StARS$model_par$Pi 
+pi_proba_df <- as.data.frame(pi_proba)
 
-  if (output == "igraph") {
+STREAM <- NR_DATA$stream
+# Assume M is a matrix and V is a vector of same length as nrow(M)
 
-    G <-  graph_from_adjacency_matrix(net, mode = "undirected", weighted = TRUE, diag = FALSE)
+# Convert to data frame
 
-    if (!is.null(node.labels)) {
-      igraph::V(G)$label <- node.labels
-    } else {
-      igraph::V(G)$label <- colnames(net)
+pi_proba_df$STREAM <- STREAM
+
+# Keep one row per unique value of V
+pi_proba_unique <- pi_proba_df %>%
+  group_by(STREAM) %>%
+  slice(1) %>%
+  ungroup()
+
+# Extract matrix part
+pi_proba_final <- as.matrix(select(pi_proba_unique, -STREAM))
+
+# Set row names as the values of V
+rownames(pi_proba_final) <- pi_proba_unique$V
+
+# Plot the heatmap
+
+pheatmap(pi_proba_final,
+         color = rev(paletteer::paletteer_c("viridis::cividis", 30)),
+         cluster_rows = FALSE,
+         cluster_cols = FALSE,
+         show_rownames = TRUE)  # show V as row names
+
+```
+
+# Cross-validation measures for PLN-network and ZIPLN-network
+```{r cross-val-folds}
+K = 5
+set.seed(1)
+folds    = split(sample(1:nrow(covar_data)), rep(1:K, length = nrow(covar_data)))
+```
+
+```{r pln-cross-val-function}
+crossval <- function(formula, data, K = 5,
+                     folds    = split(sample(1:nrow(data)),
+                                      rep(1:K, length = nrow(data))),
+                     method   = c("PLN", "ZIPLN")
+  ) {
+    method <- match.arg(method)
+    Y     <- lapply(folds, function(fold) data$Abundance[fold, ])
+    Y_hat <- lapply(folds, function(test_set) {
+      # rep(0, length(test_set))
+      train_set <- setdiff(1:nrow(data), test_set)
+      model <- do.call(method, list(formula = eval(formula), data = data, subset = train_set))
+      predict(model, data[test_set, ], type = "response")
+    })
+    TP_zero   <- unlist(lapply(1:K, function(k){sum(Y[[k]] == 0 & Y_hat[[k]] < 0.1)}))
+    TN_zero   <- unlist(lapply(1:K, function(k){sum(Y[[k]] != 0 & Y_hat[[k]] > 0.1)}))
+    FP_zero   <- unlist(lapply(1:K, function(k){sum(Y[[k]] != 0 & Y_hat[[k]] < 0.1)}))
+    FN_zero   <- unlist(lapply(1:K, function(k){sum(Y[[k]] == 0 & Y_hat[[k]] > 0.1)}))
+    precision <- unlist(lapply(1:K, function(k){TP_zero[[k]] / (TP_zero[[k]] + FP_zero[[k]])}))
+    recall    <- unlist(lapply(1:K, function(k){TP_zero[[k]] / (TP_zero[[k]] + FN_zero[[k]])}))
+    F1_score  <- unlist(lapply(1:K, function(k){TP_zero[[k]] / (TP_zero[[k]] + .5 * (FP_zero[[k]] + FN_zero) )}))
+
+    # err <- map2_dbl(Y_hat, Y, function(y_hat, y) mean((y_hat - y)^2))
+    err <- map2_dbl(Y_hat, Y, function(y_hat, y) Metrics::rmse(y_hat, y))
+    attr(err, "folds") <- folds
+    return(list("err" = err,
+                "TP_zero" = TP_zero, "TN_zero" = TN_zero,
+                "FP_zero" = FP_zero, "FN_zero" = FN_zero,
+                "precision" = precision, "recall" = recall,
+                "F1_score" = F1_score))
   }
-    ## Nice nodes
-    V.deg <- degree(G)/sum(degree(G))
-    igraph::V(G)$label.cex <- V.deg / max(V.deg) + .5
-    igraph::V(G)$size <- V.deg * 100
-    igraph::V(G)$label.color <- rgb(0, 0, .2, .8)
-    igraph::V(G)$frame.color <- NA
-    ## Nice edges
-    igraph::E(G)$color <- ifelse(igraph::E(G)$weight > 0, edge.color[1], edge.color[2])
-    if (type == "support")
-      igraph::E(G)$width <- abs(igraph::E(G)$weight)
-    else
-      igraph::E(G)$width <- 15*abs(igraph::E(G)$weight)
+```
 
-    if (remove.isolated) {
-      G <- delete.vertices(G, which(degree(G) == 0))
+## PLN-network cross-validation
+```{r}
+cv_pln_out      <- crossval(Abundance ~ 1 + PC1 + PC2 + stream + disturbance,
+                         data = NR_DATA)
+print(median(cv_pln_out$err))
+
+cv_pln_out_pc1_pc2      <- crossval(Abundance ~ 1 + PC1 + PC2,
+                                    data = NR_DATA, folds = folds)
+print(median(cv_pln_out_pc1_pc2$err))
+
+cv_pln_out_pc1_pc2_disturbance      <- crossval(Abundance ~ 1 + PC1 + PC2 + disturbance,
+                                    data = NR_DATA, folds = folds)
+print(median(cv_pln_out_pc1_pc2_disturbance$err))
+```
+
+## ZIPLN-network cross-validation
+```{r}
+cv_pln_out_ZI   <- crossval(Abundance ~ 1 + PC1 + PC2 + disturbance | stream ,
+                            data = NR_DATA, method = "ZIPLN")
+print(median(cv_pln_out_ZI$err))
+
+cv_pln_out_ZI_pc1_pc2   <- crossval(Abundance ~ 1 + PC1 + PC2 | 1 ,
+                            data = NR_DATA, method = "ZIPLN", folds = folds)
+print(median(cv_pln_out_ZI_pc1_pc2$err))
+
+cv_pln_out_ZI_pc1_pc2_disturbance   <- crossval(Abundance ~ 1 + PC1 + PC2 + disturbance | 1 ,
+                                    data = NR_DATA, method = "ZIPLN", folds = folds)
+print(median(cv_pln_out_ZI_pc1_pc2_disturbance$err))
+```
+
+# Baseline methods for comparisons
+
+```{r baselines-data-formatting}
+nb_pc <- ncol(NR_COVARIATES) - 4
+NR_PCA <-
+  NR_COVARIATES %>%
+  PCA(quali.sup = quali_ind, ncp = nb_pc, scale.unit = TRUE, graph = FALSE)
+
+NR_PCA_SCORES <- setNames(data.frame(NR_PCA$ind$coord), paste0("PC", 1:nb_pc))
+
+covariates = cbind(NR_COVARIATES, NR_PCA_SCORES)
+NR_DATA <- prepare_data(
+  counts     = NR_COUNTS,
+  covariates = covariates
+)
+covar_data <- NR_DATA %>% select(c("PC1", "PC2", "disturbance", "stream"))
+```
+## BIC
+
+```{r get-BIC-function}
+get_BIC <- function(formula, covar_data, counts_data,
+                    method = c("poisson", "log_normal")){
+  method <- match.arg(method)
+  species   <- colnames(counts_data)
+
+  models <- lapply(species, function(S){
+                   Y    <- counts_data[, S]
+                   data <- cbind(covar_data, Y)
+                   colnames(data)[[length(colnames(data))]] <- "Abundance"
+                   if(method == "poisson"){
+                     glm(eval(formula), data = data,
+                         family = poisson(link = "log"))
+                   }else{
+                     log_data <- data
+                     log_data$Abundance <- log(1 + data$Abundance)
+                     glm(eval(formula), data = log_data,
+                         family = gaussian(link = "identity"))
                    }
-    if (plot) plot(G, layout = layout)
+  })
+
+  total_logLik <- sum(sapply(models, logLik))
+  total_k <- sum(sapply(models, function(m) length(coef(m))))
+  n_obs <- length(models[[1]]$y)  # assuming all same length
+  total_BIC <- 2 * total_logLik - total_k * log(n_obs)
+  return(total_BIC)
 }
-  if (output == "corrplot") {
-    if (plot) {
-      if (ncol(net) > 100)
-        colnames(net) <- rownames(net) <- rep(" ", ncol(net))
-      G <- net
-      diag(net) <- 0
-      corrplot(as.matrix(net), method = "color", is.corr = FALSE, tl.pos = "td", cl.pos = "n", tl.cex = 0.5, type = "upper")
-    } else  {
-      G <- net
+```
+
+```{r BIC-poisson}
+
+BIC_poisson <- get_BIC("Abundance ~ PC1 + PC2 + disturbance + stream", covar_data, NR_COUNTS,
+                       method = "poisson")
+print(BIC_poisson)
+
+BIC_poisson_pcs_dist <- get_BIC("Abundance ~ PC1 + PC2 + disturbance", covar_data, NR_COUNTS,
+                       method = "poisson")
+print(BIC_poisson_pcs_dist)
+
+BIC_poisson_pcs <- get_BIC("Abundance ~ PC1 + PC2", covar_data, NR_COUNTS,
+                       method = "poisson")
+print(BIC_poisson_pcs)
+```
+
+```{r BIC-log-normal}
+
+BIC_normal <- get_BIC("Abundance ~ PC1 + PC2 + disturbance + stream", covar_data, NR_COUNTS,
+                       method = "log_normal")
+print(BIC_poisson_pcs)
+
+BIC_normal_pcs_dist <- get_BIC("Abundance ~ PC1 + PC2 + disturbance", covar_data, NR_COUNTS,
+                       method = "log_normal")
+print(BIC_normal_pcs_dist)
+
+BIC_normal_pcs <- get_BIC("Abundance ~ PC1 + PC2", covar_data, NR_COUNTS,
+                           method = "log_normal")
+print(BIC_normal_pcs)
+```
+
+## Cross-validation
+
+```{r one-species-cross-val}
+crosspredict_one_species <- function(formula, folds, data,
+                                     method = c("poisson", "log_normal")){
+  method <- match.arg(method)
+
+  res <- lapply(folds, function(test_set) {
+                         train_set <- setdiff(1:nrow(data), test_set)
+                         if(method == "poisson"){
+                           tmp   <- data[train_set,]
+                           model <- glm(eval(formula),
+                                        data = tmp,
+                                        family = poisson(link = "log"))
+                           list(Y_hat = predict(model, data[test_set, ], type = "response"),
+                                Y_fit = model$fitted.values)
+                         }else{
+                           log_data <- data
+                           log_data$Abundance <- log(1 + data$Abundance)
+                           tmp   <- log_data[train_set,]
+                           model <- glm(eval(formula),
+                                        data = tmp,
+                                        family = gaussian(link = "identity"))
+                           list(Y_hat = exp(predict(model, data[test_set, ], type = "response")) - 1,
+                                Y_fit = model$fitted.values)
                          }
                          }
-  invisible(G)
+                  )
+  Y_hat = lapply(res, function(x) x$Y_hat)
+  Y_fit = lapply(res, function(x) x$Y_fit)
+  return(list("Y_hat" = Y_hat, "Y_fit" = Y_fit))
+}
+```
+
+```{r overall-cross-val-baseline}
+
+cross_val_baseline <- function(formula, folds, covar_data, counts_data,
+                      method = c("poisson", "log_normal")){
+  method <- match.arg(method)
+
+  K <- length(folds)
+  all_Y_train <- lapply(1:5, f<- function(x) c())
+  all_Y_test     <- lapply(1:5, f<- function(x) c())
+  all_Y_fit <- lapply(1:5, f<- function(x) c())
+  all_Y_hat <- lapply(1:5, f<- function(x) c())
+  species   <- colnames(counts_data)
+
+  for(S in species){
+    Y    <- counts_data[, S]
+    data <- cbind(covar_data, Y)
+    colnames(data)[[length(colnames(data))]] <- "Abundance"
+    res <- crosspredict_one_species(formula, folds, data, method)
+    Y_hat <- res$Y_hat ; Y_fit <- res$Y_fit
+    for(k in 1:K){
+      train_set <- setdiff(1:nrow(counts_data), folds[[k]])
+      all_Y_train[[k]] <- rbind(all_Y_train[[k]], data$Abundance[train_set])
+      all_Y_test[[k]]  <- rbind(all_Y_test[[k]], data$Abundance[folds[[k]]])
+      all_Y_fit[[k]]   <- rbind(all_Y_fit[[k]], Y_fit[[k]])
+      all_Y_hat[[k]]   <- rbind(all_Y_hat[[k]], Y_hat[[k]])
+      # all_Y_hat[[k]] <- rbind(all_Y_hat[[k]], rep(0, length(folds[[k]])))
+    }
+  }
+
+  predict_err <- c() ; fit_err <- c()
+  for(k in 1:K){
+    predict_err <- c(predict_err, Metrics::rmse(all_Y_test[[k]], all_Y_hat[[k]] ))
+    fit_err     <- c(fit_err, Metrics::rmse(all_Y_train[[k]], all_Y_fit[[k]] ))
+  }
+  return(list("predict_err" = predict_err,
+              "fit_err"     = fit_err))
 }
@@
 
-## Loading & formatting data
+```{r poisson-cross-val}
+cv_poisson_glm <- cross_val_baseline("Abundance ~ PC1 + PC2 + disturbance + stream",
+                            folds, covar_data, NR_COUNTS, "poisson")
+print(median(cv_poisson_glm$fit_err))
+print(median(cv_poisson_glm$predict_err))
 
-### Regional dataset
+cv_poisson_glm_pcs <- cross_val_baseline("Abundance ~ PC1 + PC2",
+                            folds, covar_data, NR_COUNTS, "poisson")
+print(median(cv_poisson_glm_pcs$fit_err))
+print(median(cv_poisson_glm_pcs$predict_err))
 
-```{r}
-load("REGIONAL_DATASET_COUNTS.Rdata")
-load("REGIONAL_DATASET_COVARIATES.Rdata")
-```
-
-```{r}
-regional_data <- prepare_data(counts     = REGIONAL_DATASET_COUNTS,
-                              covariates = REGIONAL_DATASET_COVARIATES)
-```
-
-### Northern Range dataset
-
-```{r}
-load("NORTHERN_RANGE_DATASET_COUNTS.Rdata")
-load("NORTHERN_RANGE_DATASET_COVARIATES.Rdata")
-```
-
-```{r}
-northern_range_data <- prepare_data(counts     = NORTHERN_RANGE_DATASET_COUNTS,
-                                    covariates = NORTHERN_RANGE_DATASET_COVARIATES)
-```
-
-## Running the ZIPLN-network model
-
-We fix the seed for reproducibility.
-
-```{r}
-set.seed(1)
-```
-
-### Regional dataset
-
-```{r}
-regional_model <- ZIPLNnetwork(Abundance ~ 1 + LONGLAT, data = regional_data, control = ZIPLNnetwork_param(min_ratio = 0.001,  penalize_diagonal = FALSE), zi = "col")
+cv_poisson_glm_pcs_disturbance <- cross_val_baseline("Abundance ~ PC1 + PC2 + disturbance",
+                                folds, covar_data, NR_COUNTS, "poisson")
+print(median(cv_poisson_glm_pcs_disturbance$fit_err))
+print(median(cv_poisson_glm_pcs_disturbance$predict_err))
 
@@
 
-We select a penalty using stability selection process StARS.
-As it stability selection requires a long time to run, we memorize the corresponding penalty and later use it directly to access the model selected with StARS.
+```{r log-normal-cross-val}
+cv_log_normal  <- cross_val_baseline("Abundance ~ PC1 + PC2 + disturbance + stream",
+                            folds, covar_data, NR_COUNTS, "log_normal")
+print(median(cv_log_normal$fit_err))
+print(median(cv_log_normal$predict_err))
 
-```{r}
-# regional_model_StARS <- regional_model$getBestModel("StARS")
-regional_model_StARS <- regional_model$getModel(0.06881528)  # Avec le code git
-# regional_model_StARS <- regional_model$getModel(0.4222431) # Avec la librairie
+cv_log_normal_pcs  <- cross_val_baseline("Abundance ~ PC1 + PC2",
+                            folds, covar_data, NR_COUNTS, "log_normal")
+print(median(cv_log_normal_pcs$fit_err))
+print(median(cv_log_normal_pcs$predict_err))
+
+cv_log_normal_pcs_disturbance  <- cross_val_baseline("Abundance ~ PC1 + PC2+ disturbance",
+                                folds, covar_data, NR_COUNTS, "log_normal")
+print(median(cv_log_normal_pcs_disturbance$fit_err))
+print(median(cv_log_normal_pcs_disturbance$predict_err))
 ```
-
-### Northern Range dataset
-
-```{r}
-northern_range_model <- ZIPLNnetwork(Abundance ~ 1 + LONGLAT, data = northern_range_data, control = ZIPLNnetwork_param(min_ratio = 0.001,  penalize_diagonal = FALSE), zi = "col")
-
-```
-
-We select a penalty using stability selection process StARS.
-As it stability selection requires a long time to run, we memorize the corresponding penalty and later use it directly to access the model selected with StARS.
-
-```{r}
-
-# northern_range_model_StARS <- northern_range_model$getBestModel("StARS")
-northern_range_model_StARS <- northern_range_model$getModel(0.09219056)
-```
-
-## Resulting networks
-
-We want to place species the datasets share in the same order for readibility.
-
-```{r}
-regional_species       <- colnames(regional_data$Abundance)
-northern_range_species <- colnames(northern_range_data$Abundance)
-intersecting_species   <- intersect(northern_range_species, regional_species)
-```
-
-### Regional network
-
-Sorting species names.
-
-```{r}
-regional_sorted_species <- c(intersecting_species, regional_species[! regional_species %in% intersecting_species])
-```
-
-Extracting the association network from the model.
-
-```{r}
-regional_network        <- as.matrix(regional_model_StARS$latent_network("partial_cor"))
-regional_sorted_indices <- match(regional_sorted_species, colnames(regional_network))
-regional_network        <- regional_network[regional_sorted_indices, regional_sorted_indices] 
-```
-
-```{r}
-plot_network(regional_network,
-             type            = "partial_cor",
-             output          = "igraph",
-             edge.color      = c("#F8766D", "#00BFC4"),
-             remove.isolated = FALSE,
-             node.labels     = NULL,
-             layout          = layout_in_circle,
-             plot            = TRUE)
-```
-
-### Northern Range network
-
-Sorting species names.
-
-```{r}
-northern_range_sorted_species <- c(intersecting_species, northern_range_species[! northern_range_species %in% intersecting_species])
-```
-
-```{r}
-northern_range_network        <- as.matrix(northern_range_model_StARS$latent_network("partial_cor"))
-northern_range_sorted_indices <- match(northern_range_sorted_species, colnames(northern_range_network))
-northern_range_network        <- northern_range_network[northern_range_sorted_indices, northern_range_sorted_indices] 
-```
-
-```{r}
-plot_network(northern_range_network,
-             type            = "partial_cor",
-             output          = "igraph",
-             edge.color      = c("#F8766D", "#00BFC4"),
-             remove.isolated = FALSE,
-             node.labels     = NULL,
-             layout          = layout_in_circle,
-             plot            = TRUE)
-```

### 2025-07-23 PAPER_CODE.qmd
removed deprecated and change anablepsoides to rivulus

@@
 ```{r}
-log_abundance <- log(1 + NR_DATA$Abundance)
-log_abundance_sorted <- log_abundance[order(NR_DATA$stream), ]
+log_abundance <- log(1 + NR_COUNTS)
+log_abundance_sorted <- log_abundance[order(NR_COVARIATES$stream), ]
 
@@
          cluster_rows = FALSE, cluster_cols = FALSE,
-         annotation_row = data.frame(Stream = NR_DATA$stream),
+         annotation_row = data.frame(Stream = NR_COVARIATES$stream),
          show_rownames = FALSE)
 ```
+
 ## Covariates preparation (PCA)
@@
 # Cross-validation measures for PLN-network and ZIPLN-network
+
 ```{r cross-val-folds}
@@
 set.seed(1)
-folds    = split(sample(1:nrow(covar_data)), rep(1:K, length = nrow(covar_data)))
+folds    = split(sample(1:nrow(NR_DATA)), rep(1:K, length = nrow(NR_DATA)))
 ```
@@
 ## PLN-network cross-validation
+
 ```{r}
@@
 ## ZIPLN-network cross-validation
+
 ```{r}
@@
 ```
+
 ## BIC

### 2025-07-24 PAPER_CODE.qmd
cleaned up code for submission

@@
 ---
-title: "Species association networks unveil scale effects in community assemblage mechanisms."
-author: "Jeanne Tous, Julien Chiquet"
+title: "A JSDM with zero-inflation to increase the ecological relevance of analyses of species distribution data."
+author: "Jeanne Tous, Julien Chiquet, Amy E. Deacon, Ada Fontrodona-Eslava, Douglas F. Fraser, Anne E. Magurran"
 format: html
@@
 ```{r, loading libraries}
-# library(PLNmodels)
+library(PLNmodels)
 library(Metrics)
@@
 
-color_palette <- paletteer_c("ggthemes::Orange-Gold", n = 30)
+color_palette  <- paletteer::paletteer_c("ggthemes::Orange-Gold", n = 30)
 palette_colors <- paletteer::paletteer_c("ggthemes::Orange-Gold", n = 29)  # 29 instead of 30
@@
 nb_pc <- ncol(NR_COVARIATES) - 4
+
 NR_PCA <-
@@
 )
-covar_data <- NR_DATA %>% select(c("PC1", "PC2", "disturbance", "stream"))
+covar_data <- NR_DATA %>% dplyr::select(c("PC1", "PC2", "disturbance", "stream"))
 ```

### 2025-07-24 ZIPLN_illustration.R

@@ -1,0 +1,177 @@
+library(MASS)
+# library(PLNmodels)
+library(pheatmap)
+library(paletteer)
+
+# Function to get partial correlations from precision matrix
+partial_correlations <- function(Omega){
+  p <- nrow(Omega)
+  partial_cor <- matrix(rep(0, p*p), nrow = p)
+  diag(partial_cor) <- 1
+  for(i in 1:(p-1)){
+    for(j in (i + 1):p){
+      partial_cor[i, j] <- -Omega[i, j] / sqrt(Omega[i, i] * Omega[j, j])
+    }
+  }
+  partial_cor[lower.tri(partial_cor)] <- t(partial_cor)[lower.tri(partial_cor)]
+  return(partial_cor)
+}
+
+# Generating simulated data
+set.seed(1)
+n = 500
+p = 3
+
+X <- matrix(0.01 * as.vector(-200:299), nrow = n)
+B <- matrix(as.vector(c(2, 2, 1.7)), nrow = 1)
+
+R <- matrix(c(1, -0.05, -0.1,
+              -0.05, 1, 0.9,
+              -0.1, 0.9, 1), nrow = 3, byrow = TRUE)
+variances <- c(0.5, 0.5, 0.5)
+D <- diag(sqrt(variances))
+Sigma <- D %*% R %*% D
+Omega <- solve(Sigma)
+
+Y = matrix(rep(1, n*p), nrow=n)
+Z = mvrnorm(n, mu=matrix(rep(0, p), p, 1), Sigma=Sigma)
+for(j in 1:p){
+  for(i in 1:n){ Y[i, j] = rpois(1, exp(t(X[i,]) %*% B[,j] + Z[i,j])[1,])}
+}
+
+Yres = matrix(rep(1, n*p), nrow=n)
+for(j in 1:p){
+  for(i in 1:n){ Yres[i, j] =  rpois(1, exp( Z[i,j]))}
+}
+
+colnames(Y) <- c("A", "B", "C")
+simu_data <- prepare_data(Y, X)
+
+# Visualising the abundances
+log_Y <- log(1 + Y)
+color_palette <- paletteer_c("ggthemes::Orange-Gold", n = 30)
+palette_colors <- paletteer::paletteer_c("ggthemes::Orange-Gold", n = 29)  # 29 instead of 30
+color_palette <- c("white", palette_colors)
+max_val <- max(log_Y, na.rm = TRUE)
+
[…]

### 2025-07-24 PAPER_CODE.qmd
Update PAPER_CODE.qmd

@@
 title: "A JSDM with zero-inflation to increase the ecological relevance of analyses of species distribution data."
-author: "Jeanne Tous, Julien Chiquet, Amy E. Deacon, Ada Fontrodona-Eslava, Douglas F. Fraser, Anne E. Magurran"
+author: 
 format: html
